Laplacian envelope

### 2020-07-25 Notes/laplace.tex
heuristic envelope functions

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{?.tex} with its contents
%       or include ?.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\usepackage{mathtools}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
%
\input{sl_graphics_preamble.tex}
\graphicspath{{Figures/}}
%
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distribution
\newcommand{\pr}{\mathbf{p}}
%equilibrium distribution
\newcommand{\eq}{\pib}
%equilibrium metric
\newcommand{\Eq}{\boldsymbol{\Pi}}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\arow}{\boldsymbol{\xi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\etw}{\boldsymbol{\eta}^w}
\newcommand{\thb}{\boldsymbol{\theta}}
\newcommand{\omb}{\boldsymbol{\omega}}
\newcommand{\alb}{\boldsymbol{\alpha}}
%synapse models
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{K}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%with activity indep processes
\renewcommand{\hom}{\mathbf{H}}
\newcommand{\Mh}{\widetilde{\M}}
\newcommand{\frgh}{\widetilde{\W}^{\mathrm{F}}}
\newcommand{\ench}{\widetilde{\enc}}
%eigenvectors
\newcommand{\evr}{\mathbf{u}}
\newcommand{\evl}{\boldsymbol{\eta}}
%snr curves etc
\newcommand{\syn}{\vec{w}}
\newcommand{\synid}{\syn_\text{id}}
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\DeclareMathOperator{\NNR}{NNR}
\DeclareMathOperator{\nnr}{NNR}
\newcommand{\snrb}{\overline{\snr}}
\DeclareMathOperator{\env}{Env}
\newcommand{\rh}{\hat{r}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
%avoid double superscripts
\renewcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
% shifted process
\newcommand{\shift}[1]{\widehat{#1}}

@@

The expressions in \cref{eq:finitederivs} have the same form as for $s=0$, which allowed us to argue that the model with maximal area must have the serial topology, \cite{Lahiri2013synapse}.

The one missing ingredient is the scale invariance.
Define the scale transformation:
%
\begin{equation}\label{eq:scale}
  \lambda * \M^\nu \equiv (1-\lambda)\I + \lambda \M^\nu.
\end{equation}
%
The Laplace transform has the following scaling property
%
\begin{equation}\label{eq:laplacescale}
  \hat{A} \prn{\lambda s;\lambda * \M^\nu} = \hat{A} \prn{s; \M^\nu}.
\end{equation}
%
At $s=0$, it is scale invariant.
This means that we can take a pair of matrices that violates the ``diagonal'' constraints (second inequality of \eqref{eq:constri}) and use the scale transform \eqref{eq:scale} to construct a pair of matrices that satisfy the ``diagonal'' constraints and have the same area.
This allowed us to ignore the diagonal constraints.
We can't do that for $s>0$.

The corresponding symmetry here is scaling the shifted process from \cref{eq:lshiftpd}, which amounts to
%
\begin{equation}\label{eq:shiftscale}
  \lambda {\circledast}_{s} \M^\nu 
      \equiv (1 - \lambda) \brk{(1 + s)\I - s \onev \eq} 
      + \lambda \M^\nu.
\end{equation}
%
Then we have
%
\begin{equation}\label{eq:lshiftscale}
  \hat{A} \prn{s; \M^\nu} 
    = \hat{A} \prn{0, \shift{\M}^\nu(s)}
    = \hat{A} \prn{0, \lambda * \shift{\M}^\nu(s)}
    = \hat{A} \prn{s, \lambda {\circledast}_{s} \M^\nu}.
\end{equation}
%
The problem here is that the lower bound, \(\M^\nu_{ij} \geq 0\), is not invariant under this transformation, unlike the \(s = 0\) case.
Using this transformation to fix the upper bound can break the lower bound.

@@

@@

The relation of the Laplace transform to the shifted process from \cref{eq:lshiftpd,eq:lareashift} suggests an alternative approach: 
treat the off-diagonal components of \(\shift{\M}^\nu \) as the degrees of freedom and maximise the area instead.
The constraints on \(\shift{\M}^\nu_{ij}\) are different from \cref{eq:constri}:
%
\begin{equation}\label{eq:lshiftconstr}
  \shift{\M}^\nu_{ij} \geq s \eq_j,
    \quad \forall \; \nu, i, j : i \neq j, \qquad
  \sum_{j: j \neq i} \shift{\M}^\nu_{ij} \leq 1 + s (1 - \eq_i).
    \quad \forall \; \nu, i.
\end{equation}
%
The diagonal elements are still given by \( \shift{\M}^\nu_{ii} = 1 - \sum_{j: j \neq i} \shift{\M}^\nu_{ij} \), but now they are allowed to be negative.

@@

The derivatives, for $m \neq n$ are
%
\begin{equation}\label{eq:lshiftderivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\Mh^\nu_{mn}} = & \,
    \eq_m \prn{ \brk{\thb_m - \thb_n}
     + \prn{\mathbf{c}_m + (-1)^\nu + \lambda} \brk{\etw_m - \etw_n} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     - \sum_{ij} \mu^\nu_{ij} s f^\nu \eq_m (\fptb_{mj} - \fptb_{nj}) \eq_j
    = 0.
\end{aligned}
\end{equation}
%
We can absorb the new terms by defining
%
\begin{equation*}
\begin{aligned}
  \shift{\thb}^\nu(s) = \thb(0) - s f^\nu \fptb(0) \Eq (\boldsymbol{\mu}^\nu)\trans \onev,
  \qquad
  \shift{\alb}^\nu(s) &= (\M^\nu + (s - 1) \I - s \onev \eq) \shift{\thb}^\nu(s),
  \\
  \shift{\omb}^\nu(s) &= (\M^\nu + (s - 1) \I - s \onev \eq) \etw(0).
\end{aligned}
\end{equation*}
%
We note that
%
\begin{equation*}
  \mu^\nu_{ij} (\M^\nu_{ij} - s \eq_j) = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} (\M^\nu_{ij} - s \eq_j)
      = \mu^\nu_{ii} (1 - \M^\nu_{ii} - s (1 - \eq_i)) = \mu^\nu_{ii},
\end{equation*}
%
then we can ``solve'' for the KKT multipliers:
%
\begin{equation}\label{eq:lshiftsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eq_m \prn{ \shift{\alb}^\nu_m(s) 
    + \brk{ \mathbf{c}_m + (-1)^\nu + \lambda } \shift{\omb}^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eq_m \prn{ 
      \brk{ \shift{\thb}^\nu_n(s) - \shift{\thb}^\nu_m(s) - \shift{\alb}^\nu_m(s) }
  % \\ & \phantom{= - \eq_m \big( } 
    + \brk{ \mathbf{c}_m + (-1)^\nu + \lambda } 
        \brk{ \etw_n + \etw_m - \shift{\omb}^\nu_m(s) }
     }, \\
\end{aligned}
\end{equation}
%
although there are KKT multipliers hiding in \(\shift{\thb}^\nu(s)\) and \(\shift{\alb}^\nu(s)\).

@@

The first of these equations is solved by
%
\begin{equation}\label{eq:detagensol}
  \delta\eta_i = B\e^{\beta(i-1)} + C\e^{\beta(1-i)},
  \qquad\text{where}\qquad
  s = S(\beta) \equiv 2 \sinh^2 \prn{\frac{\beta}{2}}
    = \cosh (m\beta) - 1,
\end{equation}
%
with $B$ and $C$ determined by the boundary conditions provided by the last two equations of \eqref{eq:detahalf}:
%
\begin{equation}\label{eq:unicoeffs}
  B = -\frac{1}{s(1 + \e^{-\beta}) \cosh \prn{m \beta}},
  \qquad
  C = -\frac{1}{s(\e^{\beta} + 1) \cosh \prn{m \beta}}.
\end{equation}
%
This can be used to compute the Laplace transform
%
\begin{equation}\label{eq:unilaplace}
  A(s) = \frac{\etw_1(s)}{m}
   = \frac{1}{m}\prn{\frac{1}{s}+B+C}
   = \frac{2\sinh^2\prn{\frac{m \beta}{2}}}{m s \cosh\prn{m \beta}}
   = \frac{2 S\prn{m \beta}}{M s \prn{S\prn{m \beta} + 1}}.
\end{equation}
%

@@

>>>>

% -*- TeX -*- -*- UK -*-
% ----------------------------------------------------------------
% arXiv Paper ************************************************
%
% Subhaneil Lahiri's template
%
% Before submitting:
%    Comment out hyperref
%    Comment out showkeys
%    Replace \input{?.tex} with its contents
%       or include ?.tex in zip/tar file
%    Put this file, the .bbl file, any picture or
%       other additional files and natbib.sty
%       file in a zip/tar file
%
% **** -----------------------------------------------------------
\documentclass[12pt]{article}
%Preamble:
\usepackage{a4wide}
\input{sl_preamble.tex}
\usepackage{mathtools}
%
% >> Only for drafts! <<
\usepackage[notref,notcite]{showkeys}
%
\input{sl_graphics_preamble.tex}
\graphicspath{{Figures/}}
%
% ----------------------------------------------------------------
%\numberwithin{equation}{section}
%\renewcommand{\baselinestretch}{1.5}
% ----------------------------------------------------------------
% New commands etc.
\input{sl_definitions.tex}
\input{sl_symbols.tex}
%matrices
%vec of ones
\newcommand{\onev}{\mathbf{e}}
%mat of ones
\newcommand{\onem}{\mathbf{E}}
%Markov matrix
\newcommand{\MM}{\mathbf{Q}}
%prob distribution
\newcommand{\pr}{\mathbf{p}}
%equilibrium distribution
\newcommand{\eq}{\pib}
%equilibrium metric
\newcommand{\Eq}{\boldsymbol{\Pi}}
%first passage times
\newcommand{\fpt}{\mathbf{T}}
%off-diag first passage times
\newcommand{\fptb}{\overline{\fpt}}
%fundamental matrix
\newcommand{\fund}{\mathbf{Z}}
%other symbols
\newcommand{\Pb}{\mathbf{P}}
\newcommand{\D}{\mathbf{D}}
\newcommand{\pib}{\boldsymbol{\pi}}
\newcommand{\arow}{\boldsymbol{\xi}}
\newcommand{\Lb}{\boldsymbol{\Lambda}}
\newcommand{\lb}{\boldsymbol{\lambda}}
\newcommand{\etw}{\boldsymbol{\eta}^w}
\newcommand{\thb}{\boldsymbol{\theta}}
\newcommand{\omb}{\boldsymbol{\omega}}
\newcommand{\alb}{\boldsymbol{\alpha}}
%synapse models
\newcommand{\w}{\mathbf{w}}
\newcommand{\W}{\mathbf{W}}
\newcommand{\M}{\mathbf{M}}
\newcommand{\enc}{\mathbf{K}}
\newcommand{\frg}{\W^{\mathrm{F}}}
\newcommand{\F}{\boldsymbol{\Phi}}
%superscripts
\newcommand{\pot}{^{\text{pot}}}
\newcommand{\dep}{^{\text{dep}}}
\newcommand{\potdep}{^{\text{pot/dep}}}
%with activity indep processes
\renewcommand{\hom}{\mathbf{H}}
\newcommand{\Mh}{\widetilde{\M}}
\newcommand{\frgh}{\widetilde{\W}^{\mathrm{F}}}
\newcommand{\ench}{\widetilde{\enc}}
%eigenvectors
\newcommand{\evr}{\mathbf{u}}
\newcommand{\evl}{\boldsymbol{\eta}}
%snr curves etc
\newcommand{\syn}{\vec{w}}
\newcommand{\synid}{\syn_\text{id}}
\DeclareMathOperator{\SNR}{SNR}
\DeclareMathOperator{\snr}{SNR}
\DeclareMathOperator{\NNR}{NNR}
\DeclareMathOperator{\nnr}{NNR}
\newcommand{\snrb}{\overline{\snr}}
\DeclareMathOperator{\env}{Env}
\newcommand{\rh}{\hat{r}}
\newcommand{\CA}{\mathcal{A}}
\newcommand{\CB}{\mathcal{B}}
%avoid double superscripts
\renewcommand{\pdiff}[2]{\frac{\partial #1}{\partial #2}}
% shifted process
\newcommand{\shift}[1]{\widehat{#1}}

@@

The expressions in \cref{eq:finitederivs} have the same form as for $s=0$, which allowed us to argue that the model with maximal area must have the serial topology, \cite{Lahiri2013synapse}.
The one missing ingredient is the scale invariance.

@@

\subsection{Scale invariance}\label{sec:scale}

Define the scale transformation:
%
\begin{equation}\label{eq:scale}
  \zeta * \M \equiv \zeta \M + (1-\zeta)\I .
\end{equation}
%
The Laplace transform has the following scaling property
%
\begin{equation}\label{eq:laplacescale}
  \hat{A} \prn{\zeta s;\zeta * \M^\nu} = \hat{A} \prn{s; \M^\nu}.
\end{equation}
%
At $s = 0$, it is scale invariant.
This means that we can take a pair of matrices that violates the ``diagonal'' constraints (second inequality of \eqref{eq:constri}) and use the scale transform \eqref{eq:scale} to construct a pair of matrices that satisfy the ``diagonal'' constraints and have the same area.
This allowed us to ignore the diagonal constraints.
We can't do that for $s > 0$.

The corresponding symmetry here is scaling the shifted process from \cref{eq:lshiftpd}, which amounts to
%
\begin{equation}\label{eq:shiftscale}
  \zeta {\circledast}_{s} \M 
      \equiv \zeta \M
      + (1 - \zeta) \brk{(1 + s)\I - s \onev \eq} .
\end{equation}
%
Then we have
%
\begin{equation}\label{eq:lshiftscale}
  \hat{A} \prn{s; \M^\nu} 
    = \hat{A} \prn{0, \shift{\M}^\nu(s)}
    = \hat{A} \prn{0, \zeta * \shift{\M}^\nu(s)}
    = \hat{A} \prn{s, \zeta \circledast_s \M^\nu}.
\end{equation}
%
The problem here is that the lower bound, \(\M^\nu_{ij} \geq 0\), 
is not invariant under this transformation, unlike the \(s = 0\) case.
Using this transformation to fix the upper bound can break the lower bound.

We might ask what conditions are required so that a model is fixable?
This means that the \(\zeta \circledast_s \) orbit of the model contains some valid models.
The conditions are
%
\begin{equation*}
\begin{alignedat}{5}
  [\zeta \circledast \M^\nu]_{ij}  
    &= \zeta \M^\nu_{ij} - (1-\zeta) s \eq_j 
    &&\geq 0 &
  \quad &\implies \quad &
  \zeta &\geq \frac{s \eq_j}{\M^\nu_{ij} + s \eq_j} &
    &\forall \nu, i, j : i \neq j , \\
  \sum_{j : i \neq j} [\zeta \circledast \M^\nu]_{ij} 
    &= \zeta \lb^\nu_i - (1-\zeta) s (1-\eq_i) 
    &&\leq 1 &
  &\implies &
  \zeta &\leq \frac{1 + s (1-\eq_i)}{\lb^\nu_i + s (1-\eq_i)} &
    \quad &\forall \nu, i ,
\end{alignedat}
\end{equation*}
%
where \(\lb^\nu_i \equiv \sum_{j : i \neq j} \M^\nu_{ij}\) is the inverse mean  lifetime of state \(i\).
The model is fixable if it is possible to find a \(\zeta\) that satisfies all of the above conditions:
%
\begin{equation}\label{eq:fixable}
  \frac{s \eq_j}{\M^\nu_{ij} + s \eq_j} \leq \frac{1 + s (1-\eq_k)}{\lb^\mu_k + s (1-\eq_k)}
  \quad \implies \quad
  \frac{\lb^\mu_k - 1}{1 + s (1-\eq_k)} \leq \frac{\M^\nu_{ij}}{s \eq_j} 
  \quad \forall \mu, \nu, i, j, k : i \neq j.
\end{equation}
%
Note that when \(\M^\nu_{ij} = 0\) for some \(i,j \), this condition reduces to \(\lb^\mu_k \leq 1\), the original diagonal constraint.
Thus, relaxing the constraints from ``the model is valid'' to ``the model is fixable'' doesn't help us prove anything about the topology.

@@

The relation of the Laplace transform to the shifted process from \cref{eq:lshiftpd,eq:lareashift} suggests an alternative approach: 
treat the off-diagonal components of \(\shift{\M}^\nu \) as the degrees of freedom and maximise the area instead.
The constraints on \(\shift{\M}^\nu_{ij}\) are different from \cref{eq:constri}:
%
\begin{equation}\label{eq:lshiftconstr}
  \shift{\M}^\nu_{ij} \geq s \eq_j,
    \quad \forall \; \nu, i, j : i \neq j, \qquad
  \sum_{j: j \neq i} \shift{\M}^\nu_{ij} \leq 1 + s (1 - \eq_i).
    \quad \forall \; \nu, i.
\end{equation}
%
The diagonal elements are still given by \( \shift{\M}^\nu_{ii} = 1 - \sum_{j: j \neq i} \shift{\M}^\nu_{ij} \), but now they are allowed to be negative so long as \(\M^\nu_{ii} \geq - s (1 - \eq_i)\).

@@

The derivatives, for $m \neq n$ are
%
\begin{equation}\label{eq:lshiftderivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\Mh^\nu_{mn}} = & \,
    \eq_m \prn{ \brk{\thb_m - \thb_n}
     + \prn{\mathbf{c}_m + (-1)^\nu + \lambda} \brk{\etw_m - \etw_n} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     - \sum_{ij} \mu^\nu_{ij} s f^\nu \eq_m (\fptb_{mj} - \fptb_{nj}) \eq_j
    = 0.
\end{aligned}
\end{equation}
%
We can absorb the new terms by defining
%
\begin{equation*}
\begin{aligned}
  \shift{\thb}^\nu(s) = \thb(0) - s f^\nu \fptb(0) \Eq (\boldsymbol{\mu}^\nu)\trans \onev,
  \qquad
  \shift{\alb}^\nu(s) &= (\M^\nu + (s - 1) \I - s \onev \eq) \shift{\thb}^\nu(s),
  \\
  \shift{\omb}^\nu(s) &= (\M^\nu + (s - 1) \I - s \onev \eq) \etw(0).
\end{aligned}
\end{equation*}
%
We note that
%
\begin{equation*}
  \mu^\nu_{ij} (\M^\nu_{ij} - s \eq_j) = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} (\M^\nu_{ij} - s \eq_j)
      = \mu^\nu_{ii} (1 - \M^\nu_{ii} - s (1 - \eq_i)) = \mu^\nu_{ii},
\end{equation*}
%
then we can ``solve'' for the KKT multipliers:
%
\begin{equation}\label{eq:lshiftsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eq_m \prn{ \shift{\alb}^\nu_m(s) 
    + \brk{ \mathbf{c}_m + (-1)^\nu + \lambda } \shift{\omb}^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eq_m \prn{ 
      \brk{ \shift{\thb}^\nu_n(s) - \shift{\thb}^\nu_m(s) - \shift{\alb}^\nu_m(s) }
  % \\ & \phantom{= - \eq_m \big( } 
    + \brk{ \mathbf{c}_m + (-1)^\nu + \lambda } 
        \brk{ \etw_n - \etw_m - \shift{\omb}^\nu_m(s) }
     }, \\
\end{aligned}
\end{equation}
%
although there are KKT multipliers hiding in \(\shift{\thb}^\nu(s)\) and \(\shift{\alb}^\nu(s)\).

@@

The first of these equations is solved by
%
\begin{equation}\label{eq:detagensol}
  \delta\eta_i = B\e^{\beta(i-1)} + C\e^{\beta(1-i)},
  \qquad\text{where}\qquad
  s = S(\beta) \equiv 2 \sinh^2 \prn{\frac{\beta}{2}}
    = \cosh (\beta) - 1,
\end{equation}
%
with $B$ and $C$ determined by the boundary conditions provided by the last two equations of \eqref{eq:detahalf}:
%
\begin{equation}\label{eq:unicoeffs}
  B = -\frac{1}{s(1 + \e^{-\beta}) \cosh \prn{m \beta}},
  \qquad
  C = -\frac{1}{s(\e^{\beta} + 1) \cosh \prn{m \beta}}.
\end{equation}
%
This can be used to compute the Laplace transform
%
\begin{equation}\label{eq:unilaplace}
  A(s) = \frac{\etw_1(s)}{m}
   = \frac{1}{m}\prn{\frac{1}{s}+B+C}
   = \frac{2\sinh^2\prn{\frac{m \beta}{2}}}{m s \cosh\prn{m \beta}}
   = \frac{2 S\prn{m \beta}}{M s \prn{S\prn{m \beta} + 1}}.
\end{equation}
%

@@

This expression breaks down for \(M = 2\), as there are no \( 2 \leq i \leq m \) when \(m = 1\).
In that case, 
%
\begin{equation}\label{eq:binaryshorten}
  \hat{A}(s) = \frac{1-\varepsilon}{1-\varepsilon + s}.
\end{equation}
%
We can see that the optimal \(\varepsilon = 0\), reducing \cref{eq:binaryshorten} to \cref{eq:binaryenv}.
